---
title: "Application - Custom functions"
author: "Vanderlei J. Debastiani, André G. Coutinho, Marcos B. Carlucci, Marcus V. Cianciaruso"
date: "`r Sys.Date()`"
output:
 rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Application - Custom functions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: none
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  results = 'hold',
  fig.height = 5, 
  fig.width = 6, 
  fig.align = 'center',
  comment = "#>"
)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

This vignette illustrates the use of custom functions to calculate functional parameters of the simulated communities.

# The dataset

The dataset is detailed in the overview vignette.

```{r message=FALSE, results= "markup"}
# Load the package
require(restauraR)
# Load cerrado data
data(cerrado)
# Traits data
traits <- cerrado$traits
# Species composition in reference sites
referenceSites <- cerrado$reference
# Species composition in restoration sites
restorationSites <- cerrado$restoration
```


# Generate simulated communities

As explained in the Overview vignette, the simulation is performed using the `simulateCommunities` function, specifying input parameters. The function returns an object of class `simRest` (a list object) with basic information about the input parameters and a list of results. 


```{r}
# Set seed
set.seed(2025)
```

```{r}
# Simulation scenario
resultsSim <- simulateCommunities(traits = traits, # Species trait data
                                  # Species composition in restoration sites
                                  restComp = restorationSites, 
                                  it = 1000, # Number of simulations
                                  rich = c(10, 15), # Species richness constraints
                                  # Maximise functional diversity of all traits
                                  maxDiver = c("BT", "SLA", "Height", "Seed"),  
                                  # Constrain communities by bark thickness trait
                                  constCWM = "BT", 
                                  method = "proportions" # Sampling method
)
```

```{r, echo = FALSE}
resultsSim$call <- "simulateCommunities(...)"
```

```{r}
resultsSim
```

# Calculate functional metrics

The `computeParameters` function calculates key ecological metrics such as species richness, CWM, Rao's quadratic entropy, and taxonomic dissimilarity from reference sites. Note that the function returns an object of class `simRest` (a list) that now includes all calculated metrics.

```{r}
# Compute all functional metric for simulated communities and reference sites
resultsSim <- computeParameters(resultsSim,
                                traits = traits,
                                # Traits for Rao calculation
                                rao = c("BT", "SLA", "Height", "Seed"), 
                                # Traits for CWM calculation
                                cwm = c("BT", "SLA", "Height", "Seed"), 
                                # Reference site for ecological metric comparison
                                reference = referenceSites) 
```

```{r, echo = FALSE}
resultsSim$call <- "simulateCommunities(...)"
```

```{r}
resultsSim
```

# Custom functions

Several taxonomic, functional and phylogenetic metrics are available in the literature and in R, however, they are not included in the standard version of the `computeParameters` function. The function includes an option to calculate any metric in a customised way, allowing the user to use a wide range of metrics in the function.

The basic structure of the function includes the arguments `x`, `comm` and `traits`. The `x` argument is only used as an index for the `lapply` function. The `comm` argument specifies the simulated and the reference communities concatenated. This species composition matrix is not standardised, thus it can contain raw abundances or species proportions according to the method chosen in the simulation procedure. The `traits` argument includes original trait values according to the selected columns specified by the argument `traitsFUN`. The ellipses argument is used to pass any additional arguments directly to the function. 


```{r, eval = FALSE}
# Skeleton of FUN functions
fun <- function(x, comm, traits, ...){
  # x - mandatory argument (not used in the function)
  # comm - Raw species composition matrix
  # traits - Species traits after being selected by the traitsFUN argument
  # ... - Additional arguments to function
  res <- data.frame(metric = res$metric) # Result must be a data.frame with one or more columns
  return(res)
}
```

A small example is used to illustrate a function that calculates the CWM using non-standardised species composition. 

```{r}
funCWM <- function(x, comm, traits){
  comm <- as.matrix(comm) 
  traits <- as.matrix(traits)
  matchNames <- match(colnames(comm), rownames(traits))
  traits <- traits[matchNames,]
  res <- as.data.frame(comm%*%traits)
  return(res)
}
```

Another example can be performed using any functional metric included in the `dbFD` function of the `FD` package. In our example, we set this skeleton to calculate the functional richness. 

```{r, eval = FALSE}
funFric <- function(x, comm, traits, ...){
  res <- FD::dbFD(x = traits, a = comm, ...)
  res <- data.frame(Fric = res$FRic)
  return(res)
}
```


```{r}
# Compute all functional metric for simulated communities and reference sites
resultsSimCustom <- computeParameters(resultsSim,
                                      traits = traits,
                                      FUN = funCWM, # Custom function for CWM
                                      # Traits for custom CWM calculation
                                      traitsFUN = c("BT", "SLA", "Height", "Seed"), 
                                      reference = referenceSites)
```

```{r}
summary(resultsSimCustom)
```

```{r}
# Metric calculated for all communities simulated ...
head(resultsSimCustom$simulation$results)
```

```{r}
# ... and for the reference sites
head(resultsSimCustom$reference$results)
```


Note that the ellipse allows passing any argument directly to the custom function.


```{r, eval = FALSE}
resultsSimFric <- computeParameters(resultsSim,
                                    traits = traits,
                                    # Custom function for functional richness (Fric)
                                    FUN = funFric, 
                                    # Traits for FRic calculation
                                    traitsFUN = c("BT", "SLA", "Height", "Seed"), 
                                    # Extra argument to dbFD function related to the Fric metric
                                    stand.FRic = TRUE, 
                                    reference = referenceSites)
```


The inclusion of this flexibility allows computing taxonomic and phylogenetic metrics, in addition to other functional metrics, in a relatively easy way.
