---
title: "Application - Custom functions"
author: "Vanderlei J. Debastiani, André G. Coutinho, Marcos B. Carlucci, Marcus V. Cianciaruso"
date: "`r Sys.Date()`"
output:
 rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Application - Custom functions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: none
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  results = 'hold',
  fig.height = 5, 
  fig.width = 6, 
  fig.align = 'center',
  comment = "#>"
)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

This vignette illustrates the use of custom function to calculate functional parameters of the simulated communities.

# The dataset

The dataset is detailed in the overview vignette.

```{r message=FALSE, results= "markup"}
# Load the package
require(resbiota)
# Load cerrado data
data(cerrado.mini)
# Traits data
traits <- cerrado.mini$traits
# Species composition in reference sites
referenceSites <- cerrado.mini$reference
# Species composition in restoration sites
restorationSites <- cerrado.mini$restoration
# Species composition in supplementary sites
```


# Generate simulated communities

As explained in overview vignette, the simulation is performed using the `simulateCommunities` function, specifying input parameters. The function returns an object of class `simRest` (a list object) with basic information of input parameters and a list of results. 


```{r}
# Set seed to be able to repeat the same simulation in the future
set.seed(2025)
```

```{r}
# Simulation scenario
resultsSim <- simulateCommunities(trait = traits, # Species trait data
                                  restComp = restorationSites, # Current species composition in restoration sites
                                  it = 1000, # Number of simulations
                                  rich = c(10, 15), # Species richness constraints, min = 10, max = 15 species
                                  cwm = "BT", # Constrain communities by bark thickness trait
                                  rao = c("BT", "SLA", "Height", "Seed"),  # Maximize functional diversity of all traits
                                  method = "proportions" # Sampling method
)
resultsSim
```

# Calculate functional metrics

Again as explained in overview vignette, the `computeParameters` function calculates key ecological metrics as species richness, CWM, Rao's quadratic entropy, and taxonomic dissimilarity with reference sites. Note that the function returns an object of class `simRest` (a list) that now includes all calculated metric.

```{r}
# Compute all functional metric for simulated communities and reference sites
resultsSim <- computeParameters(resultsSim,
                                traits = traits,
                                cwm = c("BT", "SLA", "Height", "Seed"), # Traits for CWM calculation
                                rao = c("BT", "SLA", "Height", "Seed"), # Traits for Rao calculation
                                reference = referenceSites) # Reference site for ecological metric comparison
```


```{r, echo = FALSE}
resultsSim$call <- "simulateCommunities(...)"
```

```{r}
resultsSim
```


## Custom functions

Several taxonomic, functional and phylogenetic metrics are available in the literature and in R, however, not included in standard version of the `computeParameters` function. The function includes an option to calculate any metric in a customized way, allowing the user to utilize a wide range of metric in the function.

The basic structure of the function includes the arguments x, comm and traits. The `x` argument is only used as an index to the `lapply` function; `comm` arguments specify the simulated and the reference communities concatenated, this species composition matrix is not standardised, thus it can contain raw abundances or species proportions according to the method chosen in the simulation procedure; `traits` arguments include original trait values according to the selected columns specified by the argument `traitsFUN`. The ellipses arguments are used to pass any object direct to the function. 


```{r, eval = FALSE}
# Skeleton of FUN functions
fun <- function(x, comm, traits, ...){
  # x - mandatory argument, it is not used in the function
  # comm - Raw species composition
  # traits - Species traits after selected by the traitsFUN argument
  # ... - Additional arguments to function
  res <- data.frame(metric = res$metric) # Result must be a data.frame with one or more columns
  return(res)
}
```

A small example is used to illustrate a function that calculates the CWM using non-standardised species composition. 

```{r}
funCWM <- function(x, comm, traits){
  comm <- as.matrix(comm) 
  traits <- as.matrix(traits)
  matchNames <- match(colnames(comm), rownames(traits))
  traits <- traits[matchNames,]
  res <- comm%*%traits
  return(res)
}
```

Another example can be performed using any functional metric included in the 'dbFD' function. In our example, we set this skeleton to calculate the functional richness. 

```{r, eval = FALSE}
funFric <- function(x, comm, traits, ...){
  res <- FD::dbFD(x = traits, a = comm, ...)
  res <- data.frame(Fric = res$FRic)
  return(res)
}
```


```{r}
# Compute all functional metric for simulated communities and reference sites
resultsSim <- computeParameters(resultsSim,
                                traits = traits,
                                FUN = funCWM, # Custom function for CWM
                                traitsFUN = c("BT", "SLA", "Height", "Seed"), # Traits for custom CWM calculation
                                reference = referenceSites)
head(resultsSim$simulation$results)
head(resultsSim$reference$results)
```


```{r, eval = FALSE}
resultsSim <- computeParameters(resultsSim,
                                traits = traits,
                                FUN = funFric, # Custom function for functional richness (Fric)
                                traitsFUN = c("BT", "SLA", "Height", "Seed"), # Traits for FRic calculation
                                stand.FRic = TRUE, # Extra argument to dbFD function related to the Fric metric
                                reference = referenceSites)
```


The inclusion of this flexibility allows computing taxonomic and phylogenetic metrics in addition to other functional metrics in a relatively easy way.













