---
title: "Framework application"
author: "André G. Coutinho, Marcos B. Carlucci, Marcus V. Cianciaruso"
date: "`r Sys.Date()`"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Framework application}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: none
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  results = 'hold',
  fig.height = 5, 
  fig.width = 6, 
  fig.align = 'center',
  comment = "#>"
)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

The document illustrate the application of the framework to select species for ecological restoration following the trait-based approach of Coutinho, Carlucci & Cianciaruso, A framework to apply trait-based ecological restoration at large scales, Journal of Applied Ecology.

Set seed to be able to repeat the same simulation in the future. For an example or to replicate the findings in the manuscript `set.seed(2022)`.

```{r, echo = FALSE}
set.seed(2022)
```

## Step 1: Generate simulated communities and calculate parameters

### Load the package

```{r, message = FALSE}
require(resbiota)
```

### The dataset

The `cerrado` dataset contain traits to 483 species used in restoration projects in the savannas in the Cerrado biome. The dataset arranged as a list: `traits` object with traits data and `reference` object the composition of 6 reference sites.

The data frame `traits` contain four traits: `BT` (bark thickness), `SLA` (specific leaf area), `Height` (maximum height) and `Seed` (seed mass). Furthermore, the column `Available` indicates if species available on the market, column `Cost` the cost per individual and the column `Density` planting density for each species. The species names are arranged as the row names of data frame.

The matrix `reference` contain relative abundances for 118 species in the reference sites. The species names are arranged as the column names of matrix and sites names as row names. Remember to include them with relative abundances instead of raw values.

```{r}
# Load cerrado data
data(cerrado)
# Traits data
traits <- cerrado$traits
# Species composition in reference sites
referenceSites <- cerrado$reference
```

```{r, echo = FALSE}
knitr::kable(head(traits), digits = 3, row.names = TRUE, caption = "Traits data (first 6 rows only)")
knitr::kable(head(referenceSites, n = c(6, 3)), digits = 3, row.names = TRUE, caption = "Reference sites (first 3 columns only)")
```

### Species traits

In the resbiota package the selection of traits which will be used in each step is done using traits names in character type.

```{r}
allTraits <- c("BT", "SLA", "Height", "Seed")
```

### Species available

In same way, the pointing of the column which contain species available on the market is done using character type.

```{r}
# Species available
ava <- "Available"
```

### Species richness

Select the range of species richness you like to have in the random communities. In the example below, communities ranging from 20 to 70 species will be generated.

```{r}
# Set species richness
SR <- c(20, 70)
```

### Number of random communities to be generated

How large N should be? This depends mostly on (1) the number of species in the species pool and (2) the functional trait values of the species. A species pool composed by few species that are functionally similar need a lower N values than a pool composed by a several number of functionally distinct species. Therefore there is not a straight answer for the "N question" above. Depending on your findings (= the number of communities that fall within the restoration thresholds or targets) you can increase N always observing how many new selected communities this will return. You will able to check this easily in the STEP 2. But, remember that large N values will increase processing time. This is especially true if you have reference sites. We recommend N = 5000 as a minimum value, but for example propose we used only 100 communities.

```{r}
# Set number of random communities
N <- 100
```

### Community Weighted Mean

The Community Weighted Mean (CWM; Garnier et al. 2004; Lavorel et al. 2008) define which traits will be used to calculate CWM. You can use ordinal traits, but the columns in your trait data must be numeric. To use ordinal variables transform your categories into numbers. The number of traits should be less than the number of species and no missing values are tolerated. If you have species with missing trait information you should consider one of the several approaches to handling this. For some examples on this see Penone et al. 2014 (10.1111/2041-210X.12232), Taugourdeau et al. 2014 (10.1002/ece3.989), Johnson et al. 2021 (10.1111/geb.13185), Debastiani et al. 2021 (10.1016/j.ecoinf.2021.101315). You can also calculate CWM for more than one trait in object data.

```{r}
# Bark thickness
traits2cwm <- c("BT")
```

### Rao Quadratic Entropy (Rao)

The Rao Quadratic Entropy (Rao) this is a widely used measure of functional diversity that combines species abundances and functional differences among species. For some examples, see: Botta-Dukat 2005 (10.1111/j.1654-1103.2005.tb02393.x), de Bello et al. 2016 (10.1007/s00442-016-3546-0), Laughlin et al. 2018 (10.1111/2041-210X.13023). You can define either a vector indicating which traits will be used to calculate Rao or a distance matrix (a dist object). This is especially useful if your trait matrix has several types of variables (for example, continous and categorical traits). In such case you should use gower distance to create the distance matrix. Check Pavoine et al. 2009. Oikos (doi.org/10.1111/#j.1600-0706.2008.16668.x) on how to properly deal with distinct types of variables. Using an object class dist (a distance matrix) if you have reference sites to which you want to compare your solutions using functional dissimilarity, select columns in data which represent your reference sites. NAs are not allowed.

```{r}
# Using a trait vector
traits2rao <- c("SLA", "Height", "Seed")
```

### Generate simulated communities

The simulation are performed using `simulateCommunities` function, specifying input parameters. The function return an object of class `simRest` (a list object) with basic information of input parameters and a list of results.

```{r, echo = FALSE}
resSIM <- readRDS("vignettes/resSIM0.rds")
```

```{r, eval = FALSE}
resSIM <- simulateCommunities(trait = traits,
                              ava = ava,
                              it = N,
                              rich = SR,
                              cwm = traits2cwm,
                              rao = traits2rao)
```

```{r}
resSIM
```

```{r, echo = FALSE, eval = FALSE}
saveRDS(resSIM, "vignettes/resSIM0.rds")
```

### Restoration costs

If you want to calculate cost, you need to provide a vector of species cost per individual and a vector of planting density (for example, number of individuals per hectare) for each species. Ideally, density values should take into account germination rates and seedling mortality.

```{r}
# Species cost
cost <- "Cost"
# Species planting density
dens <- "Density"
```

### Calculate functional parameters

For each simulated communities, you can count species unavailable, compute richness, CWM, Rao and estimate the restoration cost. Note that this function return an object of class `simRest` (a list object) now with calculated parameters.

```{r}
# Compute all parameters for simulated communities and reference sites
resSIM <- computeParameters(resSIM,
                                    trait = traits,
                                    cwm = traits2cwm,
                                    rao = traits2rao,
                                    ava = ava,
                                    cost = cost,
                                    dens = dens,
                                    reference = referenceSites)
```

```{r}
resSIM
```

## Step 2: Select simulated communities based on restoration targets

### Explore the results

First, use the function `viewResults` to compare different parameters and define the thresholds. In the example below view the relationship between bark thickness CWM and Rao a plot similar to Figure 2A in the manuscript will be produced.

```{r}
viewResults(resSIM, xvar = "CWM_BT", yvar = "rao")
```

### Select communities

Now, define the desired thresholds to select communities. The thresholds can be set of one or more parameters, but the names must match the parameters names in `resSIM` object. For example, CWM bark thickness higher than 6 mm and Rao higher than 0.7.

```{r}
# Set thresholds
thresholds <- c("CWM_BT > 6", "rao > 0.7")
```

Now, use `selectCommunities` to filter selected communities that satisfies these thresholds.

```{r}
# Select communities
resSEL <- selectCommunities(resSIM,
                                 testsDet = thresholds)
```

```{r}
resSEL
```

See the number of selected communities.

```{r}
# See the number of selected communities
resSEL$selection$thresholds
```

### Calculating functional dissimilarity

If you have reference sites, you can calculate functional dissimilarity between selected communities and reference sites. Functional dissimilarity is a pair-wise metric between each selected community and each reference site. If you have more than a reference site, functional dissimilarity will be the average dissimilarity between each selected community and the reference sites. Functional dissimilarity is calculated with the function `discomQE` from the package `adiv` using the default formula "QE". Therefore, it corresponds to the Rao's dissimilarity between communities. This metric ranges from 0 (no dissimilarity) to 1 (maximum dissimilarity). For more details and references, please see the `adiv` package R documentation.

Now, you need to provide a trait data with traits that should be used to calculate functional dissimilarity. A distance matrix (`dist` object) is also accepted.

```{r}
traitDis <- dist(scale(traits[, c("BT", "SLA", "Seed")]))
```

Now, use function `computeDissimilarity` to calculate functional dissimilarity depending on your species trait values, some warnings may appear with the following: "In is.euclid(as.dist(dis)) : Zero distance(s)". This only means that at least a pair of species have identical trait values, resulting in zero distances. This is is the case with datasets (as ours) where trait data is missing at species level and replace by genus or family-level averages.

```{r, warning = FALSE}
resSEL <- computeDissimilarity(resSEL, traitDis)
```

```{r}
resSEL
```

Now you can check the functional dissimilarity between your selected simulated communities and reference sites. This is useful to explore the results and help to define a feasible threshold for functional dissimilarity. You can plot functional dissimilarity against any other parameter with the code below.

```{r}
viewResults(resSEL, xvar = "dissimilarity", yvar = "rao", hideref = TRUE)
```

### Additional selection

A addiction threshold can be set to select simulated communities again

```{r}
# New threshold
thresholds2 <- c("dissimilarity < 0.35")
```

Now, use `selectCommunities` to apply the second filter to select communities that satisfies these threshold.

```{r}
# Select communities
resSEL <- selectCommunities(resSEL,
                                testsDet = thresholds2)
```

```{r}
resSEL
```

```{r}
# See the number of selected communities for new threshold
resSEL$selection$thresholds
```

It is possible to explore functional dissimilarity between reference sites and N-communities assembled at random. Note this will take some time, depending on how many reference sites you have, the number of random communities and the processing power of your computer. This is necessary if you want to recreate figure 2B of the manuscript.

## Step 3: Visualize and save the results

The function `viewResults` return a `ggplot` object, thus additional function of package `ggplot2` can be used to customize the plots such titles, labels and plot themes.

### View Rao vs. CWM

A plot similar to Figure 2A in the manuscript will be produced.

```{r}
# Customized plot
viewResults(resSEL, xvar = "CWM_BT", yvar = "rao") +
  ggplot2::labs(x = "CWM Bark thickness", y = "Rao")
```

### Rao vs. functional dissimilarity

If you calculated functional dissimilarity View Rao vs. functional dissimilarity. A plot similar to Figure 2B in the manuscript will be produced but only with selected communities. If this functional dissimilarity was only calculated for selected simulated communities, only these communities must be plotted.

```{r}
# Customized plot
viewResults(resSEL, xvar = "dissimilarity", yvar = "rao", hideref = TRUE) +
  ggplot2::labs(x = "Functional dissimilarity", y = "Rao")
```

### Cost vs. species richness

View cost vs. species richness. A plot similar to Figure 1S in the manuscript will be plotted. It is also possible to calculate the "cost" of reference sites setting `hideref = FALSE`. Such "cost" will refer to the cost to plant or sown the species found in reference sites. This may be of some use to put restoration cost into perspective. Note that cost will be a function of the density parameter (for example, the number of individuals to be planted or sown per hectare considering individual establishment and or survival; if these informations are available. Our default cost is in US dollars per hectare).

```{r}
# Customized plot
viewResults(resSEL, xvar = "richness", yvar = "cost") +
  ggplot2::labs(x = "Species richness", y = "Cost (USD/ha)")
```

### Exporting your results

Community matrix for selected communities

```{r}
compositionSEL <- resSEL$selection$composition
```

Parameters calculated for selected communities

```{r}
parametersSEL <- resSEL$selection$results
```

```{r, echo = FALSE}
knitr::kable(parametersSEL, digits = 3, row.names = FALSE)
```

### Unavailable species

Check the unavailable species that were present in selected communities these are the species necessary to achieve the thresholds when selected communities have at least one species not available on the market.

```{r}
# Necessary species
unavailableSpecies <- extractResults(resSEL, 
                                     trait = traits, 
                                     ava = ava, 
                                     type = "unavailableSpecies")
```

```{r, echo = FALSE}
knitr::kable(head(unavailableSpecies$necessarySpp), row.names = FALSE, caption = "Necessary species (first 6 rows only)")
```

## References

Coutinho, A. G., Carlucci, M. B., & Cianciaruso, M. V. (2023). A framework to apply trait-based ecological restoration at large scales. Journal of Applied Ecology, 60, 1562–1571. <https://doi.org/10.1111/1365-2664.14439>
