---
title: "resbiota - Overview"
author: "Vanderlei J. Debastiani, André G. Coutinho, Marcos B. Carlucci, Marcus V. Cianciaruso"
date: "`r Sys.Date()`"
output:
 rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{resbiota - Overview}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: none
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE, 
  echo = TRUE,
  collapse = TRUE,
  results = 'hold',
  fig.height = 5, 
  fig.width = 6, 
  fig.align = 'center',
  comment = "#>"
)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

# Introduction

## The framework

The framework generates communities with distinct species compositions drawn at random from a species pool. The algorithm takes as input a data set with trait information for the species in the regional pool, that is, a list of species from which trait data are available that could be used to restore a particular ecosystem. Using species from the regional pool, the algorithm produces a set of N-simulated communities with species richness that ranges within a user-defined interval.

```{r, message = FALSE, echo = FALSE, fig.cap="Flowchart with the main steps of the framework", fig.width = 6.5}
require(DiagrammeR)
DiagrammeR::grViz(
  "digraph resbiotaFlowchart {
      # define node aesthetics
      graph[splines = ortho]
      node [fontname = Helvetica, shape = box, width = 5, height = 1, fontsize = 16, style = filled, fillcolor = whitesmoke]
      bgcolor = transparent
      
      # Nodes
      dataSpeciesTraits[label = <Species traits>, fillcolor = Gold]
      dataRestorationComp[label = <Species compostion to restoration sites>, fillcolor = PaleGoldenrod]
      dataRefenceComp[label = <Species compostion to reference sites>, fillcolor = PaleGoldenrod]
      dataSupplementaryComp[label = <Species compostion to supplementary sites>, fillcolor = PaleGoldenrod]
      simulateCommunities[label = <Set simulation input parameters<br/><i>simulateCommunities</i>>, fillcolor = SeaGreen, fontcolor = snow]
      computeParameters[label = <Compute basic parameters in each simulated community<br/><i>computeParameters</i>>, fillcolor = SeaGreen, fontcolor = snow]
      standardizeParameters[label = <Calculated parameters can be standardised<br/><i>standardizeParameters</i>>, fillcolor = MediumSeaGreen]
      computeMultifunctionality[label = <Calculate multiple restoration targets, called multifunctionality<br/><i>computeMultifunctionality</i>>, fillcolor = MediumSeaGreen]
      selectCommunities[label = <Selections of simulated communities<br/><i>selectCommunities</i>>, fillcolor = SeaGreen, fontcolor = snow]
      viewResults[label = <Basic results visualization<br/><i>viewResults</i>>, fillcolor = Goldenrod]
      viewMultifunctionality[label = <Multifunctionality visualization<br/><i>viewMultifunctionality</i>>, fillcolor = Goldenrod]
      extractResults[label = <Extract and save the results<br/><i>extractResults</i>>, fillcolor = Goldenrod]
      mergeSelection[label = <Merge selections scenarios<br/><i>mergeSelection</i>>, fillcolor = Plum]
      mergeSimulations[label = <Merge simulations scenarios<br/><i>mergeSimulations</i>>, fillcolor = Plum]
      
      # Edges
      dataSpeciesTraits -> simulateCommunities
      dataSpeciesTraits -> dataRestorationComp
      dataSpeciesTraits -> dataSupplementaryComp
      dataSpeciesTraits -> dataRefenceComp
      dataRestorationComp -> simulateCommunities
      dataRefenceComp -> computeParameters
      dataSupplementaryComp -> computeParameters
      simulateCommunities -> computeParameters
      simulateCommunities -> mergeSimulations
      mergeSimulations -> simulateCommunities
      computeParameters -> standardizeParameters
      computeParameters -> computeMultifunctionality
      computeParameters -> selectCommunities
      standardizeParameters -> selectCommunities
      computeMultifunctionality -> selectCommunities
      standardizeParameters -> computeMultifunctionality
      selectCommunities -> mergeSelection
      mergeSelection -> selectCommunities
      selectCommunities -> viewResults
      selectCommunities -> viewMultifunctionality
      selectCommunities -> extractResults
      
      # Rank
      {rank = same; computeParameters dataSupplementaryComp}
      {rank = same; simulateCommunities mergeSimulations dataRefenceComp}
      {rank = same; selectCommunities mergeSelection}
      
     }"
)
```

## Traits and additional species information

Each species in the pool can be described by functional traits that represent the ecological functions of communities, ecosystem services, or desirable characteristics for novel or restored communities. Those species can be classified into species that are available on the market and those that are not. Furthermore, the cost per individual, observed frequencies occurring in reference sites, and any additional information can be used to select species composition.

## Simulations

The species assemblages are generated in the simulateCommunities function. The species richness within simulated communities varies within a range defined by the user. Thus, it could be specified using the observed range in reference communities or desired restoration targets.

The sampling of species can be performed completely at random or selected species based on functional traits or a desired trait profile. If the cwm argument is set the species selections are constrained to specific community-weighted mean trait values, using the entire range of each trait. If the rao argument is set, the species selections are performed in order to optimize the functional diversity of a specific set of traits.

To obtain a high number of restoration solutions, all runs of the simulateCommunities function return a set of simulated communities with random species distribution, and composition based on rules set with cwm and/or rao arguments. Furthermore, if the species available is informed in the ava argument, the sample of species compositions is performed also with only available species on the market. Thus, the function returns a wide range of variability in composition to posterior selection, including compositions with available and not available species in the market.

The framework returns a community matrix indicating the proportion of individuals that need to be added for each species. If no established communities are informed, the simulated communities are set as empty communities (sites to restore start with no species, and all species must be planted for restoration). Alternatively, if established communities are set, the new species and individuals are introduced into the established communities (sites to restore can start with pre-existing species). Thus, it is possible to increase the number of ecosystem functions in ongoing restoration sites and follow an approach of adaptive management.

## Communities parameters

After the initial step of species composition simulation, several function parameters can be calculated for each community. In the resbiota package, this step is mandatory, given that those parameters are used to select a subset of simulated communities that fit the restoration target. Parameters such as species richness, Functional Diversity (FD), Community Weighted Means (CWM), the average functional dissimilarity between each simulated community, and restoration cost can be calculated. In general, those function parameters as used as a proxy for ecosystem functions relevant to the restoration under evaluation. The functions computeParameters and computeMultifunctionality can be used in this step.

The function computeParameters is mandatory and allows compute the basic parameters: richness, Community Weighted Mean, Community Weighted Variance, Rao Quadratic Entropy and functional dissimilarity. The user must define which traits will be used to calculate each metric. Additionally, if are provided the species cost per individual and information about planting density for each species, the restoration costs are estimated. When the availability of species is specified, the function counts species unavailable in each community. Furthermore, it is optional to compute functional dissimilarity between selected communities and reference sites. Functional dissimilarity is a pair-wise metric between each selected community and each reference site. If you have more than one reference site, functional dissimilarity will be the average dissimilarity between each selected community and the reference sites. Functional dissimilarity is calculated with the function discomQE from the package adiv using the default formula “QE”. Therefore, it corresponds to the Rao’s dissimilarity between communities. 

The function standardizeParameters allows the standardisation of the calculated parameters. Two methods are available: "max", which divides the values by the maximum, and "standardise", which scales the calculated parameters to zero mean and unit variance. For example, the "max" method is useful to scale dissimilarity results by scaling the values in ranges from 0 (no dissimilarity) to 1 (maximum dissimilarity).

The function computeMultifunctionality is optional and computes the matrix of multifunctionality and alpha multifunctionality. See the multifunctionality topic below.

## Select communities

The selections of simulated communities are performed using the function selectCommunities. Selections are based on simple logical tests based on thresholds. The user must define parameter thresholds, that indicate the level of functional composition above (or below) which the function is considered to be restored in each site. The thresholds can be set of one or more calculated parameters in the previous step. This is an essential step that must be based on previous studies that indicate the relationship between species traits and ecosystem functions related to the desired restoration targets. Knowledge of the regional species pool, species availability and reference sites are important to find solutions that adequately optimize restoration goals.

The logical tests are specified using a vector of character class. Each test uses the parameter name, a logical operator and the value threshold. All basic logical tests can be used, setting the default operators: \<, \<=, \>, \>=, == and !=. Each test must be in double or single quote signs. The selection of character type can be performed using different quotation marks than those used to define the entire test (e.g., single quote inside the double quote).

Two selection options can be used, individually or alternatively, based on deterministic and hierarchical selection. In the deterministic selection, all simulations that are true in the input test are returned. In this case, all parameters evaluated must be true to a community to be selected. Multiple selection steps can be performed sequentially, using more restrictive criteria (although the practical results will be the same when applied in a single step with the most restrictive criterion). In the hierarchical selection, the tests are evaluated hierarchically and only one simulation is selected. When all simulations fail in the first test, the function tries the next test. In the end, the function samples only one simulation among those which passed the all tests. In this case, an additional argument defines if the selection is performed inside a specific group. This allows, for example, the selection of one simulation for each site that will be restored. Furthermore, the hierarchical selection allows the use of two special arguments to select, the word MIN to select the minimal value and the word MAX to select the maximum value in a specific parameter.

## The reference sites

Reference sites can be included in computeParameters step. Thus, all calculated parameters also are calculated to reference sites and can be used in posterior analysis or auxiliary to the selection procedure by identifying the natural range and values of these parameters in the reference ecosystem.

The resbiota package also allows the inclusion of supplementary sites to calculate the same set of functional parameters. These supplementary sites are only used to compare proposals, for example, to evaluate the parameters in naturally regenerated species same sites.

## The outputs

The main outputs are a community matrix with species relative abundances and a data frame with the functional parameters calculated for each community. These outputs allow the user to investigate the relationship between these parameters, select communities that meet the restoration goals, and identify functionally important species that are not available on the market.

## Multifunctionality

The restoration projects can target restoring multiple ecosystem services, called multifunctionality. Thus, different restoration objectives should be assigned to distinct sites within a restoration landscape. However, when dealing with multiple functions, trade-offs are likely to arise, whereby the pursuit of one function may prevent the achievement of another. In the resbiota package, the matrix of multifunctionality can be calculated using simple logical tests in each available functional parameter. Thus, the multifunctionality of each restoration site is defined as the number of functions above (or below) a given threshold. The sum of individual tests is defined as the alpha multifunctionality metric.

The selection of simulated species composition can be performed in alpha multifunctionality metric to maximize multifunctionality between restored sites. The alpha multifunctionality also allows the selection of a simulated community when no solution satisfies all initial criteria, thus, the users can use a less restricted solution.

High values of alpha multifunctionality can be a preferable criterion for selection, meaning that, for each restoration site, the selection is performed in order to return a solution with the highest number of ecosystem functions recovered.

## Visualization

The results visualization functions are auxiliary to checking parameters in reference sites, analyzing the distribution of functional parameters in simulated communities and showing final solutions of selected species composition.

The function viewResults allows the visualization of basic results using scatter plots to show the trade-offs between several parameters, restoration costs, species richness and any calculated functional parameter. This visualization includes all simulated communities and allows includes also observed relations in reference sites.

The function viewMultifunctionality allows visualize the multifunctionality sets creating a graphical representation of the number of sites where each function has been restored, and the number of sites where combinations of functions were restored.

## Merge functions

The resbiota package includes functions to merge results set in different operations.

The function mergeSimulations concatenates simulated communities generated under different scenarios. For example, part of restoration sites can be generated from empty communities, and in another scenario, the species composition can be generated based on established communities. Thus, both scenarios can be concatenated to subsequent steps as parameter calculations.

The function mergeSelection concatenates different selection procedures. For example, if restoration sites have diverse desired restoration targets, the selection can be performed in distinct steps and then concatenated.

# Main features

The resbiota package contains an integrative set of tools to create restoration solutions to assemble communities from a regional species pool. The main functions are shown below using an illustrative example. Set the seed to be able to repeat the same simulation.

```{r, message = FALSE}
# Set seed
set.seed(2024)
# Package load
require(resbiota)
```

### The dataset

The `cerrado.mini` is a toy dataset containing only 50 sampled species of the `cerrado` dataset. The original data set includes traits of 483 species used in restoration projects in the savannas in the Cerrado biome. This toy dataset was arranged to illustrate the main features of resbiota packages, including scenarios with reference sites, restoration sites and supplementary sites.

The dataset is arranged as a list: `traits` object with traits data, `reference` object the composition of 2 reference sites, `restoration` composition of 2 restoration sites and `supplementary` composition of 2 supplementary sites. The supplementary sites are similar to the reference sites but arranged in the framework only to illustrative propose, thus, the parameters can be calculated and plotted to help the interpretation of the results.

The data frame `traits` contain four traits: `BT` (bark thickness), `SLA` (specific leaf area), `Height` (maximum height) and `Seed` (seed mass). Furthermore, the column `Available` indicates if species available on the market, column `Cost` the cost per individual and the column `Density` planting density for each species. The species names are arranged as the row names of data frame.

The composition data are matrix and contain modified relative abundances. The species names are arranged as the column names of the matrix and site names as row names.

```{r}
# Load cerrado data
data(cerrado.mini)
# Traits data
traits <- cerrado.mini$traits
# Species composition in reference sites
referenceSites <- cerrado.mini$reference
# Species composition in restoration sites
restorationSites <- cerrado.mini$restoration
# Species composition in supplementary sites
supplementarySites <- cerrado.mini$supplementary
```

```{r, echo = FALSE}
knitr::kable(head(traits), digits = 3, row.names = TRUE, caption = "Traits data (first 6 rows only)")
knitr::kable(head(referenceSites, n = c(6, 3)), digits = 3, row.names = TRUE, caption = "Reference sites (first 3 columns only)")
knitr::kable(head(restorationSites, n = c(6, 3)), digits = 3, row.names = TRUE, caption = "Restoration sites (first 3 columns only)")
knitr::kable(head(supplementarySites, n = c(6, 3)), digits = 3, row.names = TRUE, caption = "Supplementary sites (first 3 columns only)")
```

## Check species pool and reference sites

The function `checkReference` allows check species pool and functional parameters in reference sites calculating basic functional parameters and returning the summary statistic those parameters. The function return an object of class `simRestCheck` (a list object) with the descriptive statistics.

```{r}
# Initial check of the reference community
resCheck <- checkReference(reference = referenceSites,
                           trait = traits,
                           cwm = "BT", 
                           rao = c("BT", "SLA", "Height", "Seed"),
                           supplementary = supplementarySites,
                           props = c(0.75)
)
resCheck
```

## Generate simulated communities

The simulation are performed using `simulateCommunities` function, specifying input parameters. The function return an object of class `simRest` (a list object) with basic information of input parameters and a list of results.

### Simulate communities starting with empty communities

```{r}
# First scenario
scenarioA <- simulateCommunities(trait = traits, 
                                 ava = "Available", 
                                 it = 100, 
                                 rich = c(10, 15), 
                                 cwm = "BT", 
                                 rao = c("BT", "SLA", "Height", "Seed"),
                                 phi = 1,
                                 prefix = "New"
)
scenarioA
```

### Simulate communities starting with existing species

```{r}
# Second scenario
scenarioB <- simulateCommunities(trait = traits, 
                                 restComp = restorationSites,
                                 ava = "Available", 
                                 it = 100, 
                                 rich = c(10, 15), 
                                 cwm = "BT", 
                                 rao = c("BT", "SLA", "Height", "Seed"),
                                 phi = 1,
                                 prefix = "Ongoing"
)
scenarioB
```

### Merge simulation

Concatenates simulated communities generated under different scenarios.

```{r}
# Merge all scenarios
allScenarios <- mergeSimulations(scenarioA, scenarioB)
allScenarios 
```

## Calculate functional parameters

For each simulated communities, you can count species unavailable, compute richness, CWM, Rao and estimate the restoration cost.

```{r}
# Compute functional parameters
allScenarios <- computeParameters(x = allScenarios,
                                  trait = traits,
                                  ava = "Available",
                                  cwm = "BT",
                                  rao = c("SLA", "Height", "Seed"),
                                  cost = "Cost",
                                  dens = "Density",
                                  dissimilarity = c("SLA", "Height", "Seed"),
                                  reference = referenceSites,
                                  supplementary = supplementarySites)
allScenarios
```

A list can be provided for the Rao parameter to calculate diversity for multiple trait sets.

```{r}
# Compute functional parameters using list in Rao
# Make list
list2rao <- list(c("BT", "SLA", "Height", "Seed"), 
                 c("SLA", "Height", "Seed"))
# The list can be named
names(list2rao) <- c("allTraits", 
                     "chosenTraits")
list2rao
```

```{r}
# Compute functional parameters
allScenariosAlternative <- computeParameters(x = allScenarios,
                                             trait = traits,
                                             ava = "Available",
                                             cwm = "BT",
                                             rao = list2rao,
                                             cost = "Cost",
                                             dens = "Density",
                                             reference = referenceSites,
                                             supplementary = supplementarySites)
allScenariosAlternative
```

The multifunctionality also can be computed after the basic parameter calculation. Note that the alpha multifunctionality are included in the result data frame.

```{r}
# Compute multifunctionality
allScenarios <- computeMultifunctionality(x = allScenarios, 
                                 tests = c("CWM_BT > 6",
                                           "rao > 2.5"))
# All parameters calculated
allScenarios
```

## Select simulated communities

The selections are performed using the `selectCommunities` function. The filters selected communities that satisfy all input thresholds. The thresholds can be set of one or more parameters, but the names must match the parameters names in `resSIM` object.

### Sequential approach

Multiple selection steps can be performed sequentially, using more restrictive criteria in each step.

```{r}
# Global selection (first step)
targetSelect1 <- c("CWM_BT > 6", "rao > 2.5")
resSelectSim <- selectCommunities(allScenarios, 
                                  testsDet = targetSelect1)
```

```{r}
# Additional selection (second step)
targetSelect2 <- c("rao < 3", "cost < 8000")
resSelectSim <- selectCommunities(resSelectSim, 
                                  testsDet = targetSelect2)
```

```{r}
# Additional selection with multifunctionality
targetSelect3 <- c("alphamultifunctionality >= 2")
resSelectSim <- selectCommunities(resSelectSim, 
                                   testsDet = targetSelect3)
resSelectSim
```

### Distinct steps approach

The selection can also be performed in distinct steps and then concatenated. For example, if restoration sites have diverse desired restoration targets.

```{r}
# Selection Sites with the prefix "New"
target2New <- c("PREFIX == 'New'", "rao > 3")
resSelectSimPart1 <- selectCommunities(allScenarios, 
                                  testsDet = target2New)
```

```{r}
# Selection Sites with the prefix "Ongoing"
target2Ongoing <- c("PREFIX == 'Ongoing'", "rao < 1.5")
resSelectSimPart2 <- selectCommunities(allScenarios, 
                                   testsDet = target2Ongoing)
```

```{r}
# Then merge
resSelectSimMerged <- mergeSelection(resSelectSimPart1, resSelectSimPart2)
resSelectSimMerged
```

### Hierarchical approach

The selection can also be performed hierarchically in a way that only one simulation is selected.

```{r}
# Selection Sites with the prefix "New"
hierarchicalTargets <- c("CWM_BT > 6", "rao > 2.5", "cost == 'MIN'")
resSelectSimHier <- selectCommunities(allScenarios, 
                                  testsHie = hierarchicalTargets)
resSelectSimHier
```

## Visualize and explore the results

The function `viewResults` return a `ggplot` object, thus additional function of package `ggplot2` can be used to customize the plots such titles, labels and plot themes.

```{r}
# Customized plot
viewResults(allScenarios, xvar = "richness", yvar = "cost") +
  ggplot2::labs(x = "Species richness", y = "Cost (USD/ha)") +
  ggplot2::labs(title = "All simulations")
```

```{r}
# Customized plot
viewResults(resSelectSimMerged, xvar = "richness", yvar = "cost") +
  ggplot2::labs(x = "Species richness", y = "Cost (USD/ha)") +
  ggplot2::labs(title = "Selected simulations")
```

The multifunctionality results can be plotted for all simulations or selected simulations, allowing the visualisation of the global pattern of restoration goals.

```{r}
# View multifunctionality results
viewMultifunctionality(allScenarios)
```

```{r}
# View multifunctionality results
viewMultifunctionality(resSelectSimMerged)
```
