---
title: "restauraR - Overview"
author: "Vanderlei J. Debastiani, André G. Coutinho, Marcos B. Carlucci, Marcus V. Cianciaruso"
date: "`r Sys.Date()`"
output:
 rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{restauraR - Overview}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: none
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE, 
  echo = TRUE,
  collapse = TRUE,
  results = 'hold',
  fig.height = 5, 
  fig.width = 6, 
  fig.align = 'center',
  comment = "#>"
)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

# Introduction

## The framework

The framework generates communities with distinct species compositions drawn from a species pool. The algorithm takes as input a data set with trait information for the species in the regional pool, that is, a list of species from which trait data are available that could be used to restore a particular ecosystem. Using species from the regional pool, the algorithm produces a set of N-simulated communities with species richness that ranges within a user-defined interval.


```{r, message = FALSE, echo = FALSE, fig.cap="Flowchart with the main steps of the framework", fig.width = 6.5}
require(DiagrammeR)
DiagrammeR::grViz(
  "digraph restauraRFlowchart {
      # define node aesthetics
      graph[splines = ortho]
      node [fontname = Helvetica, shape = box, width = 5, height = 1, fontsize = 16, style = filled, fillcolor = whitesmoke]
      bgcolor = transparent
      
      # Nodes
      dataSpeciesTraits[label = <Species traits>, fillcolor = Gold]
      dataRestorationComp[label = <Species composition to restoration sites>, fillcolor = PaleGoldenrod]
      dataRefenceComp[label = <Species composition to reference sites>, fillcolor = PaleGoldenrod]
      dataSupplementaryComp[label = <Species composition to supplementary sites>, fillcolor = PaleGoldenrod]
      simulateCommunities[label = <Set simulation input parameters<br/><i>simulateCommunities</i>>, fillcolor = SeaGreen, fontcolor = snow]
      computeParameters[label = <Compute basic parameters in each simulated community<br/><i>computeParameters</i>>, fillcolor = SeaGreen, fontcolor = snow]
      standardizeParameters[label = <Calculated parameters can be standardised<br/><i>standardizeParameters</i>>, fillcolor = MediumSeaGreen]
      computeMultifunctionality[label = <Calculate multiple restoration targets, called multifunctionality<br/><i>computeMultifunctionality</i>>, fillcolor = MediumSeaGreen]
      selectCommunities[label = <Selections of simulated communities<br/><i>selectCommunities</i>>, fillcolor = SeaGreen, fontcolor = snow]
      viewResults[label = <Basic results visualization<br/><i>viewResults</i>>, fillcolor = Goldenrod]
      viewMultifunctionality[label = <Multifunctionality visualization<br/><i>viewMultifunctionality</i>>, fillcolor = Goldenrod]
      extractResults[label = <Extract and save the results<br/><i>extractResults</i>>, fillcolor = Goldenrod]
      mergeSelection[label = <Merge selections scenarios<br/><i>mergeSelection</i>>, fillcolor = Plum]
      mergeSimulations[label = <Merge simulations scenarios<br/><i>mergeSimulations</i>>, fillcolor = Plum]
      
      # Edges
      dataSpeciesTraits -> simulateCommunities
      dataSpeciesTraits -> dataRestorationComp
      dataSpeciesTraits -> dataSupplementaryComp
      dataSpeciesTraits -> dataRefenceComp
      dataRestorationComp -> simulateCommunities
      dataRefenceComp -> computeParameters
      dataSupplementaryComp -> computeParameters
      simulateCommunities -> computeParameters
      simulateCommunities -> mergeSimulations
      mergeSimulations -> simulateCommunities
      computeParameters -> standardizeParameters
      computeParameters -> computeMultifunctionality
      computeParameters -> selectCommunities
      standardizeParameters -> selectCommunities
      computeMultifunctionality -> selectCommunities
      standardizeParameters -> computeMultifunctionality
      selectCommunities -> mergeSelection
      mergeSelection -> selectCommunities
      selectCommunities -> viewResults
      selectCommunities -> viewMultifunctionality
      selectCommunities -> extractResults
      
      # Rank
      {rank = same; computeParameters dataSupplementaryComp}
      {rank = same; simulateCommunities mergeSimulations dataRefenceComp}
      {rank = same; selectCommunities mergeSelection}
      
     }"
)
```

## Traits and additional species information


Each species in the pool can be described by functional traits that represent the ecological functions of communities, ecosystem services, or desirable characteristics for novel or restored communities. Those species can be classified into species that are available on the market and those that are not. Furthermore, the cost per individual, observed frequencies occurring in reference sites, and any additional information can be used to select species composition.


## Simulations

The species assemblages are generated in the `simulateCommunities` function. The species richness within simulated communities varies within a range defined by the user. Thus, it could be specified using the observed range in reference communities or desired restoration targets.

The sampling of species can be performed completely at random or select species based on functional traits or a desired trait profile. If the `maxDiver` argument is specified, species selections optimise the functional diversity of a specific set of traits. Simultaneously, the `constCWM` argument can constrain the community-weighted mean trait values across the full range of each trait enabling local maximization of functional diversity within the specified community-weighted mean constraints.

To obtain a high number of restoration solutions, all runs of the `simulateCommunities` function return a set of simulated communities with random species distribution, and composition based on rules set with `maxDiver` and `constCWM` arguments. Furthermore, if the species available is informed in the `ava` argument, the sample of species compositions is performed also with only available species on the market. Thus, the function returns a wide range of variability in composition to posterior selection, including compositions with available and not available species in the market.

The framework returns a community matrix indicating the composition of individuals that need to be added for each species. If no established communities are informed, the simulated communities are set as empty communities (sites to restore start with no species, and all species must be planted for restoration). Alternatively, if established communities are set, the new species and individuals are introduced into the established communities (sites to restore can start with pre-existing species). Thus, it is possible to increase the number of ecosystem functions in ongoing restoration sites and follow an approach of adaptive management.

## Communities parameters

After the initial step of species composition simulation, several functional parameters can be calculated for each community. In the restauraR package, this step is mandatory, given that those parameters are used to select a subset of simulated communities that fit the restoration target. Parameters such as species richness, Functional Diversity (FD), Community Weighted Means (CWM), the average functional dissimilarity between each simulated community, and restoration cost can be calculated. In general, those function parameters are used as a proxy for ecosystem functions relevant to the restoration under evaluation. The functions `computeParameters` and `computeMultifunctionality` can be used in this step.

The function `computeParameters` is mandatory and computes basic parameters: richness, Community Weighted Mean, Community Weighted Variance, Rao’s Quadratic Entropy and functional dissimilarity. The user must define which traits will be used to calculate each metric. Additionally, if species cost per individual and planting density information are provided, restoration costs are estimated. When the availability of species is specified, the function counts species unavailable in each community. Furthermore, it is optional to compute functional dissimilarity between selected communities and reference sites. Functional dissimilarity is a pair-wise metric between each selected community and each reference site. If you have more than one reference site, functional dissimilarity will be the average dissimilarity between each selected community and the reference sites. Functional dissimilarity is calculated with the function `discomQE` from the package `adiv` using the default formula "QE". Therefore, it corresponds to the Rao’s dissimilarity between communities.

The function `standardiseParameters` allows the standardisation of the calculated parameters. Two methods are available: "max", which divides the values by the maximum, and "standardise", which scales the calculated parameters to zero mean and unit variance. For example, the "max" method is useful to scale dissimilarity results by scaling the values in ranges from 0 (no dissimilarity) to 1 (maximum dissimilarity).

The function `computeMultifunctionality` is optional and computes the matrix of multifunctionality and alpha multifunctionality. See the multifunctionality topic below.

## Select communities

The selections of simulated communities are performed using the function `selectCommunities`. Selections are based on simple logical tests based on thresholds. The user must define parameter thresholds that indicate the level of functional composition above (or below) which the function is considered to be restored in each site. The thresholds can be set for one or more calculated parameters in the previous step. This is an essential step that must be based on previous studies that indicate the relationship between species traits and ecosystem functions related to the desired restoration targets. Knowledge of the regional species pool, species availability and reference sites are important to find solutions that adequately optimise restoration goals.

The logical tests are specified using a vector of character class. Each test uses the parameter name, a logical operator and the value threshold. All basic logical tests can be used, setting the default operators: <, <=, >, >=, == and !=.  Each test must be in double or single quote signs. The selection of character type can be performed using different quotation marks than those used to define the entire test (e.g., single quote inside the double quote).

Two selection options can be used, individually or interchangeably: filter selection and priority selection. In the filter selection, all simulations that satisfy all the provided test conditions are returned. In this case, all parameters evaluated must be true for a community to be selected. Multiple selection steps can be performed sequentially, using more restrictive criteria (although the practical results will be the same when applied in a single step with the most restrictive criterion). In the priority selection, the tests are evaluated hierarchically following an order until a single simulation is selected. When all simulations fail in the first test, the function tries the next test. In the end, the function samples only one simulation from among those that passed all the tests. In this case, an additional argument defines if the selection is performed inside a specific group. This allows, for example, the selection of one simulation for each site that will be restored. Furthermore, the priority selection supports two special keywords: MIN to select the minimal value for a specified parameter, and MAX to select the maximum value.

## The reference sites

Reference sites can be included in the computeParameters step. Thus, all calculated parameters are also calculated for reference sites and can be used in posterior analysis or as an auxiliary to the selection procedure by identifying the natural range and values of these parameters in the reference ecosystem.

The restauraR package also allows the inclusion of supplementary sites to calculate the same set of functional parameters. These supplementary sites are only used to compare proposals, for example, to evaluate the parameters in naturally regenerated species sites.


## The outputs

The main outputs are a community matrix with species relative or raw abundances and a data frame with the functional parameters calculated for each community. These outputs allow the user to investigate the relationship among these parameters, select communities that meet the restoration goals, and identify functionally important species that are not available on the market.


## Multifunctionality

Restoration projects can target restoring multiple ecosystem services, called multifunctionality. Thus, different restoration objectives should be assigned to distinct sites within a restoration landscape. However, when dealing with multiple functions, trade-offs are likely to arise, whereby the pursuit of one function may prevent the achievement of another. In the restauraR package, the matrix of multifunctionality can be calculated using simple logical tests on each available functional parameter. Thus, the multifunctionality of each restoration site is defined as the number of functions above (or below) a given threshold. The alpha multifunctionality index is defined as the number of individual tests that are true (i.e., the count of restored functions) for a given community.

The selection of simulated species composition can be performed using the alpha multifunctionality index to maximise multifunctionality across restored sites. The alpha multifunctionality index also allows the selection of a simulated community when no solution satisfies all initial criteria, thus allowing users to select a less restricted solution.

High values of the alpha multifunctionality index can be a preferred criterion for selection, meaning that, for each restoration site, the selection is performed in order to return a solution with the highest number of ecosystem functions recovered.


## Visualization

The results visualisation functions are auxiliary to checking parameters in reference sites, analysing the distribution of functional parameters in simulated communities and showing final solutions of selected species composition.

The function `viewResults` allows the visualisation of basic results using histograms for parameter distributions, scatter plots to show the trade-offs between several parameters, restoration costs, species richness and any calculated functional parameter. This visualisation includes all simulated communities and also allows the inclusion of observed relations in reference sites.

Reference sites are highlighted as distinct points in scatter plots or as rug lines along the axes in histograms.

The function `viewMultifunctionality` allows visualisation of multifunctionality sets, creating a graphical representation of the number of sites where each function has been restored, and the number of sites where combinations of functions were restored.


## Merge functions

The restauraR package includes functions to merge result sets from different operations.

The function `mergeSimulations` concatenates simulated communities generated under different scenarios. For example, some restoration sites can be generated from empty communities, while in another scenario, the species composition can be generated based on established communities. Thus, both scenarios can be concatenated for subsequent steps, such as parameter calculations.

The function `mergeSelection` concatenates different selection procedures. For example, if restoration sites have diverse desired restoration targets, the selection can be performed in distinct steps and then concatenated.

## Multi-site approach

Selecting communities based solely on the local community may not be optimal for restoration planning across a landscape. The `optimiseSelection` function optimises the selection of simulated communities based on ecological indices calculated across multiple sites. The function generates all unique combinations of one simulated community per restoration site and calculates multi-site indices for each combination. Then, the selection of the simulated solutions can be performed on these multisite indices.

Two sets of indices are available: those derived from the multifunctionality matrix and those derived from species composition. The indices derived from the multifunctionality matrix include the landscape multifunctionality index, which represents the overall restoration success across the entire landscape; the beta multifunctionality index, which is calculated as Rao's quadratic entropy on the multifunctionality matrix, quantifying functional diversity among sites; and the average number of restored functions per site. The indices derived from species composition are calculates Rao's quadratic entropy, either with or without considering species dissimilarities, and include total diversity across all samples, average Alpha and Beta diversity among-community diversity, and Fst, which can be interpreted as local species identity excess. These species composition indices are calculated using a modified version of the `raoD` function from the `picante` package.

# Main features

The restauraR package contains an integrative set of tools to create restoration solutions by assemble communities from a regional species pool. The main functions are demonstrated below using an illustrative example. Set the seed to reproduce the same simulation.

```{r, message = FALSE}
# Set seed
set.seed(2026)
# Load the package
require(restauraR)
```

### The dataset

The `cerrado` is a toy dataset containing only 50 sampled species of the full cerrado dataset. The original data set includes traits of 483 species used in restoration projects in the savannas in the Cerrado biome. This toy dataset is designed to illustrate the main features of the restauraR package, including scenarios with reference sites and restoration sites.

The dataset is arranged as a list: the `traits` object contains traits data, the `reference` object contains the composition of 6 reference sites and the `restoration` object contains the hypothetical composition of 2 restoration sites. 

The `traits` data frame includes four traits: `BT` (bark thickness), `SLA` (specific leaf area), `Height` (maximum height) and `Seed` (seed mass). Additionally, the `Available` column indicates species availability on the market, the `Cost` column provides the cost per individual and the  `Density` column provides the planting density for each species. The species names are arranged as the row names of the data frame.

The composition data are matrix and contain modified relative abundances. The species names are arranged as the column names of the matrix and site names as row names.

```{r}
# Load cerrado data
data(cerrado)
# Traits data
traits <- cerrado$traits
# Species composition in reference sites
referenceSites <- cerrado$reference
# Species composition in restoration sites
restorationSites <- cerrado$restoration
```

```{r, echo = FALSE}
knitr::kable(head(traits), digits = 3, row.names = TRUE, caption = "Traits data (first 6 rows only)")
knitr::kable(head(referenceSites, n = c(6, 3)), digits = 3, row.names = TRUE, caption = "Reference sites (first 3 columns only)")
knitr::kable(head(restorationSites, n = c(6, 3)), digits = 3, row.names = TRUE, caption = "Restoration sites (first 3 columns only)")
```

## Check species pool and reference sites

The `checkReference` function checks the species pool and functional parameters in reference sites, calculating basic functional parameters and returning summary statistic. The function returns an object of class `simRestCheck` (a list object) with the descriptive statistics.

```{r}
# Initial check of the reference community
resCheck <- checkReference(reference = referenceSites, # Species composition in reference sites
                           traits = traits, # Species traits data
                           cwm = "BT", # Trait for cwm calculation
                           rao = c("BT", "SLA", "Height", "Seed"), # Traits for Rao calculation
                           props = 0.75) # Quantile to include in the summary
```

```{r, echo = FALSE}
resCheck$call <- "checkReference(...)"
```


```{r}
# Results
resCheck
```


## Generate simulated communities

The simulations are performed using the `simulateCommunities` function by specifying input parameters. The function returns an object of class `simRest` (a list object) with basic information on input parameters and a list of results.

### Simulate communities starting with empty communities

```{r}
# First scenario: starting with empty communities
scenarioA <- simulateCommunities(traits = traits, # Species traits data
                                 # Column name that indicates the availability of species
                                 ava = "Available", 
                                 it = 100, # Number of simulations
                                 # Species richness constraints, min = 10, max = 15
                                 rich = c(10, 15),  
                                 # Maximise functional diversity of all traits
                                 maxDiver = c("BT", "SLA", "Height", "Seed"), 
                                 # Constrain communities by bark thickness trait
                                 constCWM = "BT", 
                                 method = "proportions", # Sampling method
                                 # Weight of quadratic entropy or entropy, phi = 1 
                                 #    to maximise functional diversity
                                 phi = 1, 
                                 prefix = "New") # Simulation name
```

```{r, echo = FALSE}
scenarioA$call <- "simulateCommunities(...)"
```

```{r}
# Overview results
scenarioA
```


### Simulate communities starting with existing species

```{r}
# Second scenario: starting with existing species
scenarioB <- simulateCommunities(traits = traits, 
                                 # Species composition for restoration sites
                                 restComp = restorationSites, 
                                 ava = "Available", 
                                 it = 100, 
                                 rich = c(10, 15), 
                                 maxDiver = c("BT", "SLA", "Height", "Seed"),
                                 constCWM = "BT", 
                                 method = "proportions",
                                 phi = 1,
                                 prefix = "Ongoing") # Simulation name
```

```{r, echo = FALSE}
scenarioB$call <- "simulateCommunities(...)"
```

```{r}
# Overview results
# Note the total number of simulations
scenarioB
```


### Merge simulation

Concatenates simulated communities generated under different scenarios.

```{r}
# Merge all scenarios
allScenarios <- mergeSimulations(scenarioA, scenarioB) # Scenarios to merge
```

```{r, echo = FALSE}
allScenarios$call <- "mergeSimulations(...)"
```

```{r}
# Note the total number of simulations
allScenarios
```

## Calculate functional parameters

For each simulated community, you can count species unavailable, compute richness, CWM, Rao and estimate restoration cost.

```{r}
# Compute functional parameters
allScenarios <- computeParameters(x = allScenarios,
                                  traits = traits, 
                                  # Column name that indicates the availability of species
                                  ava = "Available", 
                                  cwm = "BT", # Traits for CWM calculation
                                  rao = c("SLA", "Height", "Seed"), # Traits for Rao calculation
                                  # Column name that indicates the cost per individual
                                  cost = "Cost", 
                                  dens = "Density", # Column name that species planting density
                                  # Traits for dissimilarity calculation
                                  dissimilarity = c("SLA", "Height", "Seed"), 
                                  # Species composition in the reference site
                                  reference = referenceSites) 
```

```{r, echo = FALSE}
allScenarios$call <- "mergeSimulations(...)"
```

```{r}
# Note that the results include the calculated parameters
allScenarios
```

A list can be provided for the Rao parameter to calculate diversity for multiple trait sets.

```{r}
# Compute functional parameters using list for Rao
# Create a list
list2rao <- list(c("BT", "SLA", "Height", "Seed"), # First set of traits for Rao calculation
                 c("SLA", "Height", "Seed")) # Second set of traits for Rao calculation
# The list can be named
names(list2rao) <- c("allTraits",
                     "chosenTraits")
list2rao
```

```{r}
# Compute functional parameters
allScenariosAlternative <- computeParameters(x = allScenarios,
                                             traits = traits,
                                             ava = "Available",
                                             cwm = "BT",
                                             # List with the two sets of trait names
                                             rao = list2rao, 
                                             cost = "Cost",
                                             dens = "Density",
                                             reference = referenceSites)
```

### Standardise parameters

After the parameters are calculated, they can be standardized.

```{r}
# Standardise parameters
allScenarios <- standardiseParameters(allScenarios,
                                      # Parameters to standardise
                                      parameters = c("dissimilarity", 
                                                     "functionalDissimilarity"), 
                                      method = "max")
```

```{r, echo = FALSE}
allScenarios$call <- "mergeSimulations(...)"
```

```{r}
# Result summaries
summary(allScenarios)
```

### Multifunctionality

The multifunctionality also can be computed after the basic parameter calculation. Note that the alpha multifunctionality index is included in the result data frame.

```{r}
# Compute multifunctionality
allScenarios <- computeMultifunctionality(x = allScenarios, 
                                          # Logical tests for multifunctionality assessment
                                          tests = c("CWM_BT > 6",
                                                    "rao > 0.25")) 
```

```{r, echo = FALSE}
allScenarios$call <- "mergeSimulations(...)"
```

```{r}
# Result summaries
# Note that the results now include the calculated multifunctionality
summary(allScenarios)
```

## Select simulated communities

The selections are performed using the `selectCommunities` function. The filters select communities that satisfy all input thresholds. Thresholds can be set of one or more parameters, but the names must match the parameters names in the `resSIM` object.

### Sequential approach

Multiple selection steps can be performed sequentially, using increasingly restrictive criteria in each step.

```{r}
# Global selection (first step)
targetSelect1 <- c("CWM_BT > 6", "rao > 0.25") # Logical tests for filter selection
resSelectSim <- selectCommunities(allScenarios, 
                                  testsFilter = targetSelect1)
```

```{r}
# Additional selection (second step)
targetSelect2 <- c("rao < 0.3", "Cost < 8000")
resSelectSim <- selectCommunities(resSelectSim, 
                                  testsFilter = targetSelect2)
```

```{r}
# Additional selection with multifunctionality
targetSelect3 <- c("alphamultifunctionality >= 2")
resSelectSim <- selectCommunities(resSelectSim, 
                                   testsFilter = targetSelect3)
```

```{r, echo = FALSE}
resSelectSim$call <- "selectCommunities(...)"
```

```{r}
# Note the total number of selected simulations
resSelectSim
```

### Distinct steps approach

Selection can also be performed in distinct steps and then concatenated. For example, if restoration sites have diverse restoration targets.

```{r}
# Selection Sites with the prefix "New"
# Note the format for selecting character type using single and double quote marks
target2New <- c("Scenario == 'New'", "rao > 0.3")
resSelectSimPart1 <- selectCommunities(allScenarios, 
                                       testsFilter = target2New)
```

```{r}
# Selection Sites with the prefix "Ongoing"
# Note the format for selecting a specific parameter range
target2Ongoing <- c("Scenario == 'Ongoing'", "rao > 0.3 & rao < 0.5 ")
resSelectSimPart2 <- selectCommunities(allScenarios, 
                                       testsFilter = target2Ongoing)
```

```{r}
# Then merge
resSelectSimMerged <- mergeSelection(resSelectSimPart1, resSelectSimPart2) # Scenarios to merge
```

```{r, echo = FALSE}
resSelectSimMerged$call <- "mergeSelection(...)"
```

```{r}
# Note the total number of selected simulations
resSelectSimMerged
```

### Priority approach

Selection can also be performed hierarchically, following a priority order in a way that only one simulation is selected.

```{r}
# Selection Sites with the prefix "New"
# Note the use of the keyword MIN to select the minimal value of the parameter
# Logical tests for priority selection
priorityTargets <- c("CWM_BT > 6", "rao > 0.25", "Cost == 'MIN'") 
resSelectSimPriority <- selectCommunities(allScenarios, 
                                          testsPriority = priorityTargets,
                                          # Column name to define site groups
                                          siteGroup = "Scenario", 
                                          # Select only one simulation by site group
                                          singleSelection = TRUE) 
```


```{r, echo = FALSE}
resSelectSimPriority$call <- "selectCommunities(...)"
```

```{r}
# Note that only one one simulation was selected per site group
resSelectSimPriority
```

## Optimise selection

Generates all unique combinations of one simulated community per restoration site and calculates multi-site indices for each combination. 

```{r}
scenarioMultisite <- optimiseSelection(resSelectSimPart2,
                                       siteGroup = "Site", # Column name to define site groups
                                       # Include reference sites in index calculations
                                       includeReference = TRUE, 
                                       traits = traits,
                                       # Traits for Rao calculation
                                       beta = c("SLA", "Height", "Seed"), 
                                       # Calculate Simpson's beta diversity
                                       calcTaxonomicBeta = TRUE) 
```

```{r, echo = FALSE}
scenarioMultisite$call <- "selectCommunities(...)"
```

```{r}
# Overview results
scenarioMultisite
```

```{r}
# Data.frame with multi-site indices
head(scenarioMultisite$selection$multisite$results)
```

### Select communities on multisite 

```{r}
scenarioMultisiteSel <- selectCommunities(scenarioMultisite,
                                       # Logical tests for multi-site selection
                                       testsMultisite = "alphaFunctionalDiversity == 'MAX'") 
```

```{r, echo = FALSE}
scenarioMultisiteSel$call <- "selectCommunities(...)"
```

```{r}
# Result summaries
summary(scenarioMultisiteSel)
```

## Visualise and explore the results

The `viewResults` function returns a `ggplot` object, so additional functions from the `ggplot2` package can be used to customise the plots, such as titles, labels and plot themes.

```{r}
# Histogram of calculated parameter
# Note that parameter values in the reference sites are shown as rug lines along the axis
viewResults(allScenarios, 
            xvar = "richness",
            showReference = TRUE) + # Include reference sites
  ggplot2::labs(x = "Species richness") + # Customised labels
  ggplot2::labs(title = "All simulations") # Customised title
```


```{r}
# Scatter plot of all simulations
viewResults(allScenarios, 
            xvar = "richness", 
            yvar = "Cost",
            showReference = TRUE) +
  ggplot2::labs(x = "Species richness", y = "Cost (USD/ha)") +
  ggplot2::labs(title = "All simulations")
```

```{r}
# Scatter plot of selected simulations
viewResults(resSelectSimMerged, 
            xvar = "richness", 
            yvar = "Cost",
            showReference = TRUE) +
  ggplot2::labs(x = "Species richness", y = "Cost (USD/ha)") +
  ggplot2::labs(title = "Selected simulations")
```

```{r}
# Multisite results
viewResults(scenarioMultisite, 
             xvar = "betaDiversity", 
             showMultisite = TRUE) + # Plot results from the multisite analysis
   ggplot2::labs(x = "Beta diversity", y = "Count") +
   ggplot2::labs(title = "Multisite results")
```

Multifunctionality results can be plotted for all simulations or selected simulations, allowing visualisation of the global pattern of restoration goals.

```{r}
# View multifunctionality results
viewMultifunctionality(allScenarios)
```

```{r}
# View multifunctionality results
viewMultifunctionality(resSelectSimMerged)
```

## Extract results

The function `extractResults` can be used to easily extract specific components from simulation results.

```{r}
# Extract calculated parameters of selected simulation
extractResults(scenarioMultisiteSel, 
               type = "simResults")
```

```{r}
# Species composition of selected simulation
extractResults(scenarioMultisiteSel, 
               type = "simComposition", 
               dbFormat = TRUE) # Reorganise the data in three columns format
```

