---
title: "Vignette Title2"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title2}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: none
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE, 
  echo = TRUE,
  collapse = TRUE,
  results = 'hold',
  fig.height = 5, 
  fig.width = 6, 
  fig.align = 'center',
  comment = "#>"
)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

The document illustrate the main features of CCC package. Set seed to be able to repeat the same simulation.

```{r, echo = FALSE}
set.seed(2024)
```

## Package load

```{r, message = FALSE}
require(CCC)
```

## The dataset

The `cerrado.mini` is a toy dataset containing only 50 sampled species of the `cerrado` dataset. The original data set includes traits of 483 species used in restoration projects in the savannas in the Cerrado biome. This toy dataset was arranged to illustrate the main features of CCC packages, including scenarios with reference sites, restoration sites and supplementary sites. 

The dataset is arranged as a list: `traits` object with traits data, `reference` object the composition of 2 reference sites, `restoration` composition of 2 restoration sites and `supplementary` composition of 2 supplementary sites. The supplementary sites are similar to the reference sites but arranged in the framework only to illustrative propose, thus, the parameters can be calculated and plotted to help the interpretation of the results. 

The data frame `traits` contain four traits: `BT` (bark thickness), `SLA` (specific leaf area), `Height` (maximum height) and `Seed` (seed mass). Furthermore, the column `Available` indicates if species available on the market, column `Cost` the cost per individual and the column `Density` planting density for each species. The species names are arranged as the row names of data frame.

The composition data are matrix and contain modified relative abundances. The species names are arranged as the column names of the matrix and site names as row names.

```{r}
# Load cerrado data
data(cerrado.mini)
# Traits data
traits <- cerrado.mini$traits
# Species composition in reference sites
referenceSites <- cerrado.mini$reference
# Species composition in restoration sites
restorationSites <- cerrado.mini$restoration
# Species composition in supplementary sites
supplementarySites <- cerrado.mini$supplementary
```

```{r, echo = FALSE}
knitr::kable(head(traits), digits = 3, row.names = TRUE, caption = "Traits data (first 6 rows only)")
knitr::kable(head(referenceSites, n = c(6, 3)), digits = 3, row.names = TRUE, caption = "Reference sites (first 3 columns only)")
knitr::kable(head(restorationSites, n = c(6, 3)), digits = 3, row.names = TRUE, caption = "Restoration sites (first 3 columns only)")
knitr::kable(head(supplementarySites, n = c(6, 3)), digits = 3, row.names = TRUE, caption = "Supplementary sites (first 3 columns only)")
```

## Generate simulated communities

The simulation are performed using `simulateCommunities` function, specifying input parameters. The function return an object of class `simRest` (a list object) with basic information of input parameters and a list of results.

### Simulate communities starting without species

```{r}
scenarioA <- simulateCommunities(trait = traits, 
                               ava = "Available", 
                               it = 100, 
                               rich = c(10, 15), 
                               cwm = "BT", 
                               rao = c("BT", "SLA", "Height", "Seed"),
                               phi = 1,
                               prefix = "New"
)
scenarioA
scenarioA$simulation$group
scenarioA$simulation$composition
```


### Simulate communities starting with existing species

```{r}
scenarioB <- simulateCommunities(trait = traits, 
                              restComp = restorationSites,
                               ava = "Available", 
                               it = 100, 
                               rich = c(10, 15), 
                               cwm = "BT", 
                               rao = c("BT", "SLA", "Height", "Seed"),
                               phi = 1,
                              max_add = 1,
                               prefix = "Ongoing"
)
scenarioB
scenarioB$simulation$group
scenarioB$simulation$composition
```

### Merge simulation

```{r}
allScenarios <- mergeSimulations(scenarioA, scenarioB)
allScenarios 
allScenarios$simulation$group
allScenarios$simulation$composition
```

## Calculate functional parameters

For each simulated communities, you can count species unavailable, compute richness, CWM, Rao and estimate the restoration cost. 

```{r}
# Compute functional parameters
allScenarios <- computeParameters(x = allScenarios,
                              trait = traits,
                    ava = "Available",
                    cwm = "BT",
                    rao = c("SLA", "Height", "Seed"),
                    cost = "Cost",
                    dens = "Density",
                    reference = referenceSites,
                    supplementary = supplementarySites)
allScenarios
# Compute dissimilarity
allScenarios <- computeDissimilarity(x = allScenarios, 
                                 trait = traits)
allScenarios
# Compute multifunctionality
allScenarios <- computeMultifunctionality(x = allScenarios, 
                                 tests = c("CWM_BT > 8",
                                           "rao > 2.5"))
allScenarios
```

# AQUI

### Restoration costs

If you want to calculate cost, you need to provide a vector of species cost per individual and a vector of planting density (for example, number of individuals per hectare) for each species. 

```{r}
# Species cost
cost <- "Cost"
# Species planting density
dens <- "Density"
```



## Select simulated communities

Now, use `selectCommunities` to filter selected communities that satisfies these thresholds.
Now, define the desired thresholds to select communities. The thresholds can be set of one or more parameters, but the names must match the parameters names in `resSIM` object. For example, CWM bark thickness higher than 6 mm and Rao higher than 0.7.

```{r}
### Global selection (first step) ----
targetSelect1 <- c("CWM_LMA > 0.08", "rao > 2.9", "CWM_Resprouter > 0.5")
resSelectSim <- selectCommunities(resParAllSIM, 
                                  tests = targetSelect1)
resSelectSim
class(resSelectSim)
resSelectSim$selection$results %>% dim
resSelectSim$selection$results
resSelectSim$selection$N
resSelectSim$selection$thresholds

### Additional selection (second step) ----
targetSelect2 <- c("CWM_LMA > 0.08", "rao < 3", "CWM_Resprouter > 0.5")
resSelectSim <- selectCommunities(resSelectSim, 
                                  tests = targetSelect2)
resSelectSim
str(resSelectSim,1)
resSelectSim$selection$group %>% dim
resSelectSim$selection$composition
resSelectSim$selection$results
resSelectSim$selection$N
resSelectSim$selection$thresholds

### Global selection with dissimilarity and/or multifunctionality ----
# targetSelect3 <- c("multifunctionality >= 2", "dissimilarity < 0.1")
targetSelect3 <- c("alphamultifunctionality >= 2")
resSelectSim3 <- selectCommunities(resParAllSIM, 
                                   tests = targetSelect3)
resSelectSim3
resSelectSim3$selection$results
resSelectSim3$selection$N
resSelectSim3$selection$thresholds

## Merge selections ----
### Selections ----
resParAllSIM$simulation$results
targetSelect4 <- c("PREFIX == 'Ongoing'", "restGroup == 'nonEdge'", "rao > 3.6")
resSelectSimPart1 <- selectCommunities(resParAllSIM, 
                                  tests = targetSelect4)
resSelectSimPart1
resSelectSimPart1$selection$results
resSelectSimPart1$selection$N

resParAllSIM$simulation$group
targetSelect5 <- c("PREFIX == 'Ongoing'", "restGroup == 'edge'", "rao > 3.6")
resSelectSimPart2 <- selectCommunities(resParAllSIM, 
                                   tests = targetSelect5)
resSelectSimPart2
resSelectSimPart2$selection$results
resSelectSimPart2$selection$N

### Merge ----
resSelectSimMerged <- mergeSelection(resParAllSIM, resSelectSimPart2)
resSelectSimMerged
resSelectSimMerged$selection$group %>% dim
resSelectSimMerged$selection$composition %>% dim
resSelectSimMerged$selection$results %>% dim
```

### Explore the results

First, use the function `viewResults` to compare different parameters and define the thresholds. In the example below view the relationship between bark thickness CWM and Rao a plot similar to Figure 2A in the manuscript will be produced.

```{r}
viewResults(resSIM, xvar = "CWM_BT", yvar = "rao")
```

### Select communities


```{r}
# Set thresholds
thresholds <- c("CWM_BT > 6", "rao > 0.7")
```



```{r}
# Select communities
resSEL <- selectCommunities(resSIM,
                                 tests = thresholds)
```

```{r}
resSEL
```

See the number of selected communities for each threshold (and for all them).

```{r}
# See the number of selected communities
resSEL$selection$N
```

### Calculating functional dissimilarity

If you have reference sites, you can calculate functional dissimilarity between selected communities and reference sites. Functional dissimilarity is a pair-wise metric between each selected community and each reference site. If you have more than a reference site, functional dissimilarity will be the average dissimilarity between each selected community and the reference sites. Functional dissimilarity is calculated with the function `discomQE` from the package `adiv` using the default formula "QE". Therefore, it corresponds to the Rao's dissimilarity between communities. This metric ranges from 0 (no dissimilarity) to 1 (maximum dissimilarity). For more details and references, please see the `adiv` package R documentation.

Now, you need to provide a trait data with traits that should be used to calculate functional dissimilarity. A distance matrix (`dist` object) is also accepted.

```{r}
traitDis <- dist(scale(traits[, c("BT", "SLA", "Seed")]))
```

Now, use function `computeDissimilarity` to calculate functional dissimilarity depending on your species trait values, some warnings may appear with the following: "In is.euclid(as.dist(dis)) : Zero distance(s)". This only means that at least a pair of species have identical trait values, resulting in zero distances. This is is the case with datasets (as ours) where trait data is missing at species level and replace by genus or family-level averages.

```{r, echo = FALSE}
resSEL <- readRDS("vignettes/resSEL.rds")
```


```{r, warning = FALSE, eval = FALSE}
resSEL <- computeDissimilarity(resSEL, traitDis)
```

```{r}
resSEL
```


```{r, echo = FALSE, eval = FALSE}
saveRDS(resSEL, "vignettes/resSEL.rds")
```

Now you can check the functional dissimilarity between your selected simulated communities and reference sites. This is useful to explore the results and help to define a feasible threshold for functional dissimilarity. You can plot functional dissimilarity against any other parameter with the code below.

```{r}
viewResults(resSEL, xvar = "dissimilarity", yvar = "rao", hideref = TRUE)
```

### Additional selection

A addiction threshold can be set to select simulated communities again

```{r}
# New threshold
thresholds2 <- c("dissimilarity < 0.35")
```

Now, use `selectCommunities` to apply the second filter to select communities that satisfies these threshold.

```{r}
# Select communities
resSEL <- selectCommunities(resSEL,
                                tests = thresholds2)
```

```{r}
resSEL
```

```{r}
# See the number of selected communities for new threshold
resSEL$selection$N
```

It is possible to explore functional dissimilarity between reference sites and N-communities assembled at random. Note this will take some time, depending on how many reference sites you have, the number of random communities and the processing power of your computer. This is necessary if you want to recreate figure 2B of the manuscript.

## Step 3: Visualize and save the results

The function `viewResults` return a `ggplot` object, thus additional function of package `ggplot2` can be used to customize the plots such titles, labels and plot themes.

### View Rao vs. CWM

A plot similar to Figure 2A in the manuscript will be produced.

```{r}
# Customized plot
viewResults(resSEL, xvar = "CWM_BT", yvar = "rao") +
  ggplot2::labs(x = "CWM Bark thickness", y = "Rao")
```

### Rao vs. functional dissimilarity

If you calculated functional dissimilarity View Rao vs. functional dissimilarity. A plot similar to Figure 2B in the manuscript will be produced but only with selected communities. If this functional dissimilarity was only calculated for selected simulated communities, only these communities must be plotted.

```{r}
# Customized plot
viewResults(resSEL, xvar = "dissimilarity", yvar = "rao", hideref = TRUE) +
  ggplot2::labs(x = "Functional dissimilarity", y = "Rao")
```

### Cost vs. species richness

View cost vs. species richness. A plot similar to Figure 1S in the manuscript will be plotted. It is also possible to calculate the "cost" of reference sites setting `hideref = FALSE`. Such "cost" will refer to the cost to plant or sown the species found in reference sites. This may be of some use to put restoration cost into perspective. Note that cost will be a function of the density parameter (for example, the number of individuals to be planted or sown per hectare considering individual establishment and or survival; if these informations are available. Our default cost is in US dollars per hectare).

```{r}
# Customized plot
viewResults(resSEL, xvar = "richness", yvar = "cost") +
  ggplot2::labs(x = "Species richness", y = "Cost (USD/ha)")
```

### Exporting your results

Community matrix for selected communities

```{r}
compositionSEL <- resSEL$selection$composition
```

Parameters calculated for selected communities

```{r}
parametersSEL <- resSEL$selection$results
```

```{r, echo = FALSE}
knitr::kable(parametersSEL, digits = 3, row.names = FALSE)
```

### Unavailable species

Check the unavailable species that were present in selected communities these are the species necessary to achieve the thresholds when selected communities have at least one species not available on the market.

```{r}
# Necessary species
unavailableSpecies <- extractResults(resSEL, trait = traits, ava = ava, type = "unavailableSpecies")
```

```{r, echo = FALSE}
knitr::kable(head(unavailableSpecies$necessarySpp), row.names = FALSE, caption = "Necessary species (first 6 rows only)")
```

```{r, echo = FALSE, eval = FALSE}


# CCC Example ----

?checkReference

## Initial check of the reference community ----
head(dados$trait)
resCheck <- checkReference(reference = dados$ref,
                           trait = dados$trait,
                           cwm = dados$cwm,
                           rao = dados$cwm,
                           supplementary = dados$supplementary,
                           props = c(0.75)
)
resCheck
str(resCheck, 1)
resCheck$pool$results
resCheck$pool$summary
resCheck$reference$results
resCheck$reference$summary
resCheck$supplementary$results
resCheck$supplementary$summary

## Exploratory report (html)
createReport(dados$trait, props = c(0.1, 0.5, 0.75))

## Calculate parameters ----
### Basic parameters ----

#### Merged simulation ----
resParAllSIM <- computeParameters(allSIM, 
                                    trait = dados$trait, 
                                    cwm = dados$cwm,
                                    rao = dados$cwm,
                                    # cwv = dados$cwm,
                                    ava = dados$ava,
                                    ref = dados$ref[1:10,],
                                    supplementary = dados$ref[11:19,]
)
resParAllSIM
class(resParAllSIM)
str(resParAllSIM, 2)

#### Non merged simulation ----
resParSIM0 <- computeParameters(resSIM0, 
                   trait = dados$trait, 
                   cwm = dados$cwm,
                   rao = dados$cwm,
                   cwv = dados$cwm,
                   ava = dados$ava,
                   ref = dados$ref[1:10,],
                   supplementary = dados$ref[11:19,]
)
resParSIM0
class(resParSIM0)
str(resParSIM0, 2)
resParAllSIM$simulation$composition
resParAllSIM$simulation$group
resParAllSIM$simulation$results

resParAllSIM$reference$composition
resParAllSIM$reference$results

resParAllSIM$supplementary$composition
resParAllSIM$supplementary$results


### Multifunctionality ----
resCheck$reference$summary
head(resParAllSIM$simulation$results)

target <- c("CWM_LMA > 0.105", "rao > 2.9", "CWM_Resprouter < 0.76")
# CONFERIR ----
# QUANDO NAO TEM ESPACOS ELES FICAM NOS NOMES DAS COLUNAS 
target <- c("richness > 10", "unavailable < 10", "CWM_LMA > 0.9", "CWM_Dur_flowering > 4", "rao > 2.9", "CWM_Resprouter < 0.76")

# TAREFA ----
# conferir se colunas foram encontradas
# calcular multi das referencia tambem
resParAllSIM <- computeMultifunctionality(resParAllSIM,
                            tests = target)
resParAllSIM
head(resParAllSIM$simulation$results)
head(resParAllSIM$simulation$multifunctionality)
head(resParAllSIM$simulation$multifunctionality)
dim(resParAllSIM$simulation$multifunctionality)
resParAllSIM$simulation$results

### Dissimilarity ----
resParAllSIM <- computeDissimilarity(resParAllSIM, 
                                       dados$trait[,1:2])
resParAllSIM$simulation$results

# TAREFAS ----
# incluir dissimilaridade entre comunidades selecionadas

# Select communities ----



## Extras ----

### Selection then compute dissimilarity and multifunctionality ----

#### First selection ----
targetSelect6 <- c("rao > 5")
resParSelectExtra <- selectCommunities(resParSIM0, 
                                   tests = targetSelect6)
resParSelectExtra
resParSelectExtra$selection$N
resParSelectExtra$selection$results

#### Dissimilarity ----
resParSelectExtra <- computeDissimilarity(resParSelectExtra, dados$trait[,1:2])
resParSelectExtra
resParSelectExtra$selection$results

#### Multifunctionality ----
targetMulti <- c("CWM_Height > 50", "CWM_Resprouter > 0.05")
resParSelectExtra <- computeMultifunctionality(resParSelectExtra, tests = targetMulti)
resParSelectExtra
resParSelectExtra$selection$multifunctionality
resParSelectExtra$selection$results

#### Additional selection (second step) ----
targetSelect7 <- c("alphamultifunctionality >= 2")
resParSelectExtra <- selectCommunities(resParSelectExtra, tests = targetSelect7)
resParSelectExtra
resParSelectExtra$selection$results
resParSelectExtra$reference$results
resParSelectExtra$supplementary$results

## Plots ----

### View results
viewResults(resParAllSIM, "CWM_LMA", "richness")
# viewResults(resParAllSIM, "CWM_LMA", "restGroup")
viewResults(resSelectSim, "CWM_LMA", "richness")
# viewResults(resSelectSimMerged, "CWM_LMA", "richness")

### View multifunctionality results
resParAllSIM

# Ver de mudar as cores das barras mo grafico
head(resParAllSIM$simulation$multifunctionality)

colSums(resParAllSIM$simulation$multifunctionality)
viewMultifunctionality(resParAllSIM)
viewMultifunctionality(resParAllSIM, min_degree = 3, max_degree = 6, mode = "exclusive_intersection")
viewMultifunctionality(resParAllSIM, min_degree = 3, max_degree = 6, mode = "inclusive_intersection")


resParAllSIM_TESTE <- computeParameters(allSIM, 
                                        trait = dados$trait, 
                                        # cwm = dados$cwm,
                                        rao = dados$cwm,
                                        # cwv = dados$cwm,
                                        # ava = dados$ava,
                                        ref = dados$ref[1:10,],
                                        supplementary = dados$ref[11:19,]
)
resParAllSIM_TESTE$simulation$results


resParAllSIM_TESTE2 <- computeParameters(allSIM, 
                                         trait = dados$trait, 
                                         # cwm = dados$cwm,
                                         rao = list(dados$cwm, c("LMA", "Zooc")),
                                         # cwv = dados$cwm,
                                         # ava = dados$ava,
                                         ref = dados$ref[1:10,],
                                         supplementary = dados$ref[11:19,]
)
resParAllSIM_TESTE2$simulation$results

listarao <- list(dados$cwm, c("LMA", "Zooc"))
names(listarao) <- c("Tudo", "LMAZooc")
listarao
resParAllSIM_TESTE3 <- computeParameters(resSIM1, 
                                         trait = dados$trait, 
                                         # cwm = dados$cwm,
                                         rao = listarao,
                                         # cwv = dados$cwm,
                                         # ava = dados$ava,
                                         ref = dados$ref[1:10,],
                                         supplementary = dados$ref[11:19,]
)
resParAllSIM_TESTE3$simulation$results
# END ----
```